---
title: "Class 10: Halloween Mini-Project"
author: "Sophia Wang (A16838155)"
format: gfm
---

```{r}
candy_file <- "candy-data.csv"

candy = read.csv(candy_file, row.names=1)
head(candy)
```

>Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```

>Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

```{r}
candy["Twix", ]$winpercent
```
>Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

Sour patch kids.
```{r}
candy["Sour Patch Kids","winpercent"]
```


>Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat",]$winpercent
```


>Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars",]$winpercent
```
```{r}
library(dplyr)
candy%>%filter(rownames(candy)=="Tootsie Roll Snack Bars")|>
  select(winpercent)
```
>Q. Find fruity candy with a winpercent above 50%

```{r}
candy%>%filter(fruity==1)%>%filter(winpercent>50)
```
```{r}
top.candy <- candy[candy$winpercent>50,]
top.candy[top.candy$fruity==1,]
```


To get a quick insight into a new dataset some folks like using the skier package and its `skim()` function
```{r}
library("skimr")
skim(candy)
```

>Q6.Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

Looks like the `winpercent` variable or column is meaured on a different scale than everthing else! I will need to scale my data before doing any analysis like PCA etc.

>Q7. What do you think a zero and one represent for the candy$chocolate column?

A one means "yes that is a chocolate candy" and zero means "no that is not a chocolate candy".

>Q8. Plot a histogram of winpercent values

We can do this a few ways, e.g. the "base" R `hist()` function or with `ggplot()`
```{r}
hist(candy$winpercent,breaks=10)
library(ggplot2)
ggplot(candy)+
  aes(winpercent)+
  geom_histogram(binwidth = 8)+
  theme_bw()
```

>Q9. Is the distribution of winpercent values symmetrical?

No

>Q10. Is the center of the distribution above or below 50%?

```{r}
summary(candy$winpercent)
```

Below 50%

>Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
fruit.candy <- candy|>
  filter(fruity==1)

summary(fruit.candy$winpercent)

#choco.candy <- candy[as.logical(candy$chocolate),]$winpercent
choco.candy <- candy|>
  filter(chocolate==1)
summary(choco.candy$winpercent)
```
On average chocolate candy is higher ranked than fruit candy

>Q12. Is this difference statistically significant?

```{r}
t.test(choco.candy$winpercent,fruit.candy$winpercnet)
```

Yes, the difference is significant.

>Q13. What are the five least liked candy types in this set?

```{r}
play <- c("a","d","c")
sort(play)
order(play)
play[order(play)]
```
```{r}
head(candy[order(candy$winpercent),],5)
```

>Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
head(candy[order(candy$winpercent,decreasing=T),],5)
```

>Q15. Make a first barplot of candy ranking based on winpercent values.
>Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

Let's do a barplot of winpercent values
```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col()

```

```{r}

ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col()

```

I want a more custom color scheme where I cna see both chocolate and bar and fruity etc. all from the one plot. To do thiss we can roll our own color vector...
```{r}
#Place holder color vector
mycols <- rep("black",nrow(candy))
mycols[as.logical(candy$chocolate)] <- "chocolate"
mycols[as.logical(candy$bar)] <- "brown"
mycols[as.logical(candy$fruity)] <- "pink"
mycols[row.names(candy)=="Sour Patch Kids"] <- "blue"
mycols
```

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=mycols)

```

>Q17. What is the worst ranked chocolate candy?

Sixlets

>Q18. What is the best ranked fruity candy?

Starburst

```{r}
mycols[as.logical(candy$fruity)]<-"purple"
mycols[row.names(candy)=="Sour Patch Kids"] <- "blue"
```

Add label,ggrepel
```{r}
library(ggrepel)
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=mycols) + 
  geom_text_repel(col=mycols, size=3.3, max.overlaps = 8)
```

>Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Reese's Miniatures

>Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
tail(candy[order(candy$pricepercent),0],5)
```
Hershey's Special Dark, Mr Good Bar, Ring pop, Nik L Nip, Nestle Smarties. The least popular one is Nik L Nip.

```{r}
library(corrplot)
cij <- cor(candy)
corrplot(cij,diag=F)
```
>Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Chocolate and fruity.

>Q23. Similarly, what two variables are most positively correlated?

Chocolate and winpercent.
```{r}
cij
```

```{r}
pca <- prcomp(candy, scale=T)
summary(pca)

```
```{r}
plot(pca$x[,1],pca$x[,2],col=mycols,pch=16)
```
How do the original variables (columns) contribute to the new PCs. I will look at PC1 first here
```{r}
loadings <-as.data.frame(pca$rotation) 
ggplot(loadings)+
  aes(PC1,reorder(rownames(loadings),PC1),fill=PC1)+
  geom_col()
```

```{r}
my_data <- cbind(candy, pca$x[,1:3])

p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=mycols)#+
  #geom_text()

p
```


```{r}
p + geom_text_repel(size=3.3, col=mycols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (purple), other (black)",
       caption="Data from 538")
```

```{r}
par(mar=c(8,4,2,2))

barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```
>Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

Fruity, hard, and pluribus are picked up strongly by PC1 in the positive direction. Yes these make sense because these three variables are shown to be correlated in the corrplot.
```{r}
pca$rotation[,1]
```

