---
title: "Class 15"
author: "Sophia Wang (A16838155)"
format: pdf
---
##Background
Pertussis, aka whooping cough, is a highly infections lung disease cuased by the bacteria *B. Pertussis*.

[CDC data](The CDC tracks pertussis cases numbers per year. Lets have a closer look at this data:
https://www.cdc.gov/pertussis/php/surveillance/pertussis-cases-by-year.html?CDC_AAref_Val=https://www.cdc.gov/pertussis/surv-reporting/cases-by-year.html)

We will use the **datapasta** R package to scrape this data into R.

>Q1. With the help of the R “addin” package datapasta assign the CDC pertussis case number data to a data frame called cdc and use ggplot to make a plot of cases numbers over time.

```{r}
cdc<-data.frame(
                               year = c(1922L,1923L,1924L,1925L,1926L,
                                        1927L,1928L,1929L,1930L,1931L,1932L,
                                        1933L,1934L,1935L,1936L,1937L,1938L,
                                        1939L,1940L,1941L,1942L,1943L,
                                        1944L,1945L,1946L,1947L,1948L,1949L,
                                        1950L,1951L,1952L,1953L,1954L,1955L,
                                        1956L,1957L,1958L,1959L,1960L,
                                        1961L,1962L,1963L,1964L,1965L,1966L,
                                        1967L,1968L,1969L,1970L,1971L,1972L,
                                        1973L,1974L,1975L,1976L,1977L,1978L,
                                        1979L,1980L,1981L,1982L,1983L,
                                        1984L,1985L,1986L,1987L,1988L,1989L,
                                        1990L,1991L,1992L,1993L,1994L,1995L,
                                        1996L,1997L,1998L,1999L,2000L,
                                        2001L,2002L,2003L,2004L,2005L,2006L,
                                        2007L,2008L,2009L,2010L,2011L,2012L,
                                        2013L,2014L,2015L,2016L,2017L,2018L,
                                        2019L,2020L,2021L,2022L,2024L),
       cases = c(107473,164191,165418,152003,
                                        202210,181411,161799,197371,166914,
                                        172559,215343,179135,265269,180518,
                                        147237,214652,227319,103188,183866,
                                        222202,191383,191890,109873,133792,
                                        109860,156517,74715,69479,120718,68687,
                                        45030,37129,60886,62786,31732,28295,
                                        32148,40005,14809,11468,17749,
                                        17135,13005,6799,7717,9718,4810,3285,
                                        4249,3036,3287,1759,2402,1738,
                                        1010,2177,2063,1623,1730,1248,1895,
                                        2463,2276,3589,4195,2823,3450,4157,
                                        4570,2719,4083,6586,4617,5137,
                                        7796,6564,7405,7298,7867,7580,9771,
                                        11647,25827,25616,15632,10454,13278,
                                        16858,27550,18719,48277,28639,
                                        32971,20762,17972,18975,15609,18617,
                                        6124,2116,3044,23544)
     )
```

```{r}
library(ggplot2)
baseplot<-ggplot(cdc)+
  aes(year, cases)+
  geom_point()+
  geom_line()
baseplot
```
Add some landmark developments as annotation to our plot. We include the first whole-cell (wP) vaccine roll-out in 1946.

Lets add the switch to acellular vaccine (aP) in 1996.

>Q2. Using the ggplot geom_vline() function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice?

We went from around 200,000 cases pre wP vaccine to around 1000 cases in 1976. The US switched to the aP vaccine in 1996. We start to see a big increase in 2004 to around 26,000 cases.
The number of cases decreased after the introduction of wP vaccine, while the number of cases increased after the introduction of aP vaccine.

```{r}
baseplot+
  geom_vline(xintercept=1946,col="blue")+
  geom_vline(xintercept=1996,col="red")+
  geom_vline(xintercept=2020,col="grey")+
  geom_vline(xintercept = 2003,col="green")
```

>Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

The number of cases increased after the introductin of the aP vaccine. One possible explanation for the observed trend is the vaccine conspiracy going on at that time that caused people to avoid getting vaccinated.

The is a ~10year lag from aP roll out to increasing case numbers. This holds true of other countires like Japan, UK, etc.

**Key question**: Why does aP vaccine induced immunity wane faster than that of the wP vaccine.

##CMI-PB

The CMI-PB (computational models of immunity pertusis boost) makes available lost of data about the immune response to Pertussis boost vaccination.

Critically, it tracks wP and aP individuals over time to see their immune response changes.

CMI-PB make all their data freely available via JSON format tables from their database.

Lets read the first one of these tabels:
```{r}
library(jsonlite)
subject <- read_json("http://cmi-pb.org/api/v5/subject",simplifyVector = T)
head(subject)
```
>Q. How many subjects are there in this dataset?

```{r}
nrow(subject)
```


>Q4. How many aP and wP infancy vaccinated subjects are in the dataset?

```{r}
sum(subject$infancy_vac=="aP")
sum(subject$infancy_vac=="wP")
table(subject$infancy_vac)
```

>Q5. How many Male and Female subjects/patients are in the dataset?

```{r}
table(subject$biological_sex)
```

>Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?

```{r}
table(subject$race,subject$biological_sex)
```

>Q7. Using this approach determine (i) the average age of wP individuals, (ii) the average age of aP individuals; and (iii) are they significantly different?

Average age of wP individuals is 35.5 years, and the average age of aP individuals is 26.8 years. They are significantly different.

```{r}
library(lubridate)
subject$age <- today() - ymd(subject$year_of_birth)
time_length(mean(subject[subject$infancy_vac=="wP","age"]),"years")
time_length(mean(subject[subject$infancy_vac=="aP","age"]),"years")
```

>Q8. Determine the age of all individuals at time of boost?

```{r}
int <- ymd(subject$date_of_boost) - ymd(subject$year_of_birth)
age_at_boost <- time_length(int, "year")
head(age_at_boost)
```


>Q. Does this do a good job of representing the US populus?

No. 

Lets get more data from CMI-PB, this time about the specimens collected.

```{r}
specimen <- read_json("http://cmi-pb.org/api/v5/specimen",simplifyVector = T)
head(specimen)
```
Now we can join (merge) these two tables `subject` and `specimen` to make one new `metat` table with the combined data.

>Q9. Complete the code to join specimen and subject tables to make a new merged data frame containing all specimen records along with their associated subject details:


```{r}
library(dplyr)
meta<-inner_join(subject,specimen)
head(meta)
```
Now read an "experiment data" table CMI-PB


>Q10. Now using the same procedure join meta with titer data so we can further analyze this data in terms of time of visit aP/wP, male/female etc.

```{r}
abdata <- read_json("http://cmi-pb.org/api/v5/plasma_ab_titer",simplifyVector = T)
head(abdata)
```
One more join to do of `meta` and `abdata` to assoiate all the metadata about the individual and their race, biological sex and infincy vaccination status together with Antibody levels...

```{r}
ab<-inner_join(abdata, meta)
head(ab)
```

>How many Ab measurements do we have?

```{r}
nrow(ab)
```

>Q11. How many specimens (i.e. entries in abdata) do we have for each isotype?

How many isotypes
```{r}
table(ab$isotype)
```

How many antigens?
```{r}
table(ab$antigen)
```

>Q12. What are the different $dataset values in abdata and what do you notice about the number of rows for the most “recent” dataset?

```{r}
table(ab$dataset)
```


Lets focus in IgG-one of the main antibody types responsive to bacteria or viral infections

```{r}
igg <- filter(ab,isotype=="IgG")
head(igg)
```

Make a first plot of MFI (Mean Fluorescence Intensity-a measure of much is detected) for each antigen.

```{r}
ggplot(igg)+
  aes(MFI_normalised, antigen)+
  geom_boxplot()
```
Lets color by aP/wP infancy_vac
```{r}
ggplot(igg)+
  aes(MFI_normalised, antigen,col=infancy_vac)+
  geom_boxplot()
```
```{r}
ggplot(igg)+
  aes(MFI_normalised, antigen)+
  geom_boxplot()+
  facet_wrap(~infancy_vac)
```
>Q13. Complete the following code to make a summary boxplot of Ab titer levels (MFI) for all antigens:

```{r}
ggplot(igg)+
  aes(MFI_normalised, antigen)+
  geom_boxplot()+
  facet_wrap(~visit)
```
>Q14. What antigens show differences in the level of IgG antibody titers recognizing them over time? Why these and not others?

PT, PRN, FHA and FIM2/3. These proteins are thought to participate in the bacterial infection process and they are the components included in the acellular vaccine.

```{r}
table(igg$visit)
```

Looks like we dont have data yet for all subjects in terms of visits 8 onwards. So lets exclude these.

```{r}
igg_7 <- filter(igg,visit %in% 1:7)
table(igg_7$visit)
```

```{r}
ggplot(igg_7)+
  aes(MFI_normalised, antigen,col=infancy_vac)+
  geom_boxplot()+
  facet_wrap(~visit,ncol=2)
```
Lets try a different plot. First focus on one antigen, start with PT (Pertussis Toxin) and plot visit or time on the x-axis and MFI_normalized on the y-axis.


```{r}
abdata.21 <- ab %>% filter(dataset == "2021_dataset")

abdata.21 %>% 
  filter(isotype == "IgG",  antigen == "PT") %>%
  ggplot() +
    aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,
        group=subject_id) +
    geom_point() +
    geom_line() +
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
  labs(title="2021 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)")
```

Lets finish here for today. We are beginning to see some interesting differences between aP and wP individuals. There is likely lots of other interesting thing to find in this dataset...







