---
title: "Section 4: Homework"
author: "Sophia Wang (A16838155)"
format: gfm
---
##Section 4: Population Scale Analysis
One sample is obviously not enough to know what is happening in a population. You are
interested in assessing genetic differences on a population scale. So, you processed about
~230 samples and did the normalization on a genome level. Now, you want to find whether
there is any association of the 4 asthma-associated SNPs (rs8067378...) on ORMDL3
expression.

How many samples do we have?
```{r}
expr<- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```


```{r}
nrow(expr)
table(expr$geno)
```


```{r}
library(ggplot2)
```

Lets make a box plot

```{r}
boxplot <- ggplot(expr)+
  aes(geno,exp,fill=geno)+
  geom_boxplot(notch=T)
boxplot
```

>Q13: Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes. 

```{r}
table(expr$geno)
```

```{r}
library(dplyr)
```


```{r}
medians <- expr %>%
  group_by(geno) %>%
  summarize(median_exp = median(exp))
medians
```
>Q14: Generate a boxplot with a box per genotype, what could you infer from the relative
expression value between A/A and G/G displayed in this plot? Does the SNP effect the
expression of ORMDL3? 

People with G|G geneotype have a lower expression level of ORMDL3 compared to people with A|A genotype.
SNP do affect the expression of ORMDL3. 
